---
title: "Read Counts to DGE, Part III"
author: "Friederike Duendar, ABC"
date: "September 7, 2015"
output: pdf_document
---



```{r settingUp, warning=FALSE, echo=FALSE, message=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE,cache=TRUE)
library(DESeq2)
library(magrittr)
load("ReadCountsToDGE.RData")
```

We need to ensure that the fold change will be calculated using the WT as the base line.
`DESeq` used the levels of the condition to determine the order of the comparison.

```{r DESeqPrep_relevel}
str(DESeq.ds$condition)
DESeq.ds$condition <- relevel(DESeq.ds$condition, ref="WT")
str(DESeq.ds$condition)
```

#### Brief excursion: paired analysis design

```{r eval = FALSE,}
# for paired experimental designs, an additive model is recommended
# rather than an interaction term which would be indicated using a ":"
￼design(dds) <- formula(~ patient + tissue) # last variable will automatically be used for the contrast later on
```

Running the analysis.

```{r}
DESeq.ds <- DESeq(DESeq.ds)

# DESeq() is the same as running:
DESeq.ds <- estimateSizeFactors(DESeq.ds) # sequencing depth normalization between the samples
DESeq.ds <- estimateDispersions(DESeq.ds) # gene-wise dispersion estimates across all samples
DESeq.ds <- nbinomWaldTest(DESeq.ds) # this fits a negative binomial GLM and applies Wald statistics to each gene
```

Extract the base means across samples, log2 fold changes, standard errors, 
test statistics, p-values and adjusted p-values for every gene using `results()`.

```{r}
DESeq.ds <- DESeq(DESeq.ds)
DGE.results <- results(DESeq.ds,
                       independentFiltering = TRUE,
                       alpha = 0.05)
str(DGE.results)
head(DGE.results)
summary(DGE.results)

# the DESeqResult object can basically be handled like a data.frame
head(DGE.results)
table(DGE.results$padj < 0.05)
```
NAs in the padj column (but values in both log2FC and pvalue) are indicative of that gene being filtered out by the independent filtering [because it was very lowly expressed].

The __MA-plot__ provides a global view of the differential genes, with the log2 fold
change on the y-axis over the mean of normalized counts.

Genes that pass the significance threshold (adjusted p.value \textless 0.05) are colored in red.

```{r MAplots}
par(mfrow=c(1,2))
plotMA(DGE.results, alpha = 0.05, 
       main = "Test: p.adj.value < 0.05", ylim = c(-4,4))
```

A adj. p-value histogram:

```{r adjpvalueHistogram}
par(mfrow=c(1,2))
hist(DGE.results$padj, 
     col="grey", border="white", xlab="", ylab="", main="frequencies of adj. p-values\n(all genes)")
```

A sorted results table:

```{r}
DGE.results.sorted <- DGE.results[order(DGE.results$padj),]
head(DGE.results.sorted)
```

Plotting counts for single genes (seq. depth normalized, log2-transformed)

```{r singleGenePlot}
par(mfrow=c(1,2))
plotCounts(DESeq.ds, gene="YAL056W", normalized = TRUE)  
plotCounts(DESeq.ds, gene=which.max(DGE.results$padj), main = "Max. p.adj.")
```

`plotCounts` simply uses `counts(dds, normalized = TRUE) + 0.5`.

You can also use `pcaExplorer` for individual gene plots of `rlog` values.

A heatmap of the genes that show differential expression with adjusted p-value \textless 0.05 :

```{r heatmaps}
# identify genes with the desired adjusted p-value cut-off
DGEgenes <- rownames(subset(DGE.results.sorted, padj < 0.05))

# extract rlog-transformed values into a matrix
hm.mat_DGEgenes <- rlog.norm.counts[DGEgenes,]

library(NMF)
par(mfrow=c(1,2))
# heatmap of DEG sorted by p.adjust
aheatmap(hm.mat_DGEgenes, scale="none", Rowv = NA, Colv = NA)
aheatmap(hm.mat_DGEgenes, scale="row", Rowv = NA, Colv = NA)

# combine the heatmap with hierarchical clustering (Euclidean distance, average linkage)
aheatmap(head(hm.mat_DGEgenes,50), scale="none",
         Rowv = TRUE, Colv = TRUE,
         distfun = "euclidean", hclustfun = "average",
         main = "top 50 DEG (p.adjust)")
aheatmap(head(hm.mat_DGEgenes,50), scale="row",
         Rowv = TRUE, Colv = TRUE,
         distfun = "euclidean", hclustfun = "average",
         main = "top 50 DEG ((p.adjust))")
```


## Number 1 sanity check: is SNF2 affected in the SNF2 mutant yeast samples?

To find this out, we need to retrieve the gene names and match them to the ORF 
IDs that we've used so far.
[http://www.bioconductor.org/packages/3.1/data/annotation/](http://www.bioconductor.org/packages/3.1/data/annotation/)
lists annotation packages that are available within R through bioconductor.

We will go with `org.Sc.sgd.db`.

```{r checkForSNF2}
#source("http://bioconductor.org/biocLite.R")
#biocLite("org.Sc.sgd.db")
library(org.Sc.sgd.db) # org.Hs.eg.db, org.Mm.eg.db

# list keytypes that are available to query the annotation data base
keytypes(org.Sc.sgd.db)

# list columns that can be retrieved from the annotation data base
columns(org.Sc.sgd.db)

# make a batch retrieval for all DE genes
DGEgenes <- rownames(subset(DGE.results.sorted, padj < 0.05))

anno.DGE <- select(org.Sc.sgd.db,
               keys = DGEgenes, # rownames
               keytype="ORF", # our rownames are ORF identifiers
               columns=c("SGD","GENENAME")) # what to return

# check whether SNF2 pops up among the top downregulated genes
#head(anno.DGE[match(DGEgenes, anno.DGE$ORF),])
head(anno.DGE)

```

To get a feeling for how the difference between WT and snf2 ko looks like for a
housekeeping gene, let's repeat the exercise.

```{r singleGenePlot_noDGE, echo = FALSE}
# retrieve annotation for all genes
anno.all <- select(org.Sc.sgd.db,
               keys = rownames(DGE.results.sorted), keytype="ORF",
               columns=c("SGD","GENENAME","ENTREZID","ENSEMBL"))
# get actin ORF ID
orf.actin <- anno.all$ORF[which(grepl("ACT", anno.all$GENENAME))]
# plot
library(grDevices) # for italicizing gene name
par(mfrow=c(1,2))
plotCounts(dds = DESeq.ds,
           gene = "YOR290C",
           normalized = TRUE, transform = FALSE,
           main = expression(atop("Expression of "*italic("snf2"), "(YOR290C)")))
#substitute(paste("Expression of ", italic("snf2"),"(YOR290C)")))
plotCounts(dds = DESeq.ds,
           gene = orf.actin,
           transform = FALSE, normalized = TRUE,
           main = expression(atop("Expression of "*italic("actin"), "(YFL039C)")))
# expression(atop("Expression of "*italic("actin"), b), list(b=orf.actin) ) )

```

Export the log2FC, p-values etc. into a text file:

```{r, eval=FALSE}
out.df <- merge(as.data.frame(DGE.results), anno, by.x = "row.names", by.y = "ORF")

# this will return all values for all genes
write.table(out.df, file = "DESeq2results_WT-vs-SNF2.txt", sep = "\t", 
            quote = FALSE, row.names = FALSE)
```

How can you export only those with an adjusted p-value \textless 0.05?

```{r echo=FALSE, eval=FALSE}
write.table(subset(out.df, padj < 0.05), file = "DESeq2results_WT-vs-SNF2.txt",
            sep = "\t", quote = FALSE, row.names = FALSE)
```

## GO term enrichment

Couldn't get it to work with yeast yet.

```{r goseq_yeast, eval = FALSE}
# extract Entrez IDs for the genes of interest
##DGEgenes.entrez <- subset(anno.all, ORF %in% DGEgenes, select = "ENTREZID")[,1] 
##limma::goana(DGEgenes.entrez, geneid = DGEgenes.entrez)
#DGEgenes.GO <- select(org.Sc.sgd.db, keys = DGEgenes, keytype = "ORF", columns = c("ORF", "GENENAME", "GO"))
```

Manual work-around for mouse:

```{r goseq_mm, eval = FALSE}
#biocLite("org.Mm.eg.db")
#biocLite("goseq")
#biocLite("biomaRt")
library(biomaRt)
library(org.Mm.eg.db)
library(goseq)
library(geneLenDataBase)

# gene list from http://amp.pharm.mssm.edu/Enrichr/#
# saved as genesGOtest.txt
gns <- read.table("genesGOtest.txt", stringsAsFactors = FALSE)
# retrieve the ENSEMBL symbols for the gene names
anno.mm <- select(org.Mm.eg.db,
               keys = gns$V1, 
               keytype="SYMBOL", # our rownames are ORF identifiers
               columns=c("ENSEMBL","SYMBOL","GENENAME")) # what to return

# in addition, retrieve all possible mouse gene names
# because goseq wants to know about the universe of all genes
mart <- useMart(biomart = "ensembl", dataset = "mmusculus_gene_ensembl")
mm_all <- getBM(attributes = c("ensembl_gene_id"),  mart = mart)

# get length data using goseq's data base of gene lengths
# first, load the data from the geneLenDataBase package
data(list = "mm9.ensGene.LENGTH", package = "geneLenDataBase")

# split the length values per ENSEMBL ID
len_data = split(get("mm9.ensGene.LENGTH")$Length, 
            get("mm9.ensGene.LENGTH")$Gene)

# get the gene names from the length data base
gn_names = get("mm9.ensGene.LENGTH")$Gene

# calculate the median (and other) lengths for the 
# different transcripts
len = data.frame(Gene = names(len_data), 
                 Median = sapply(len_data,median), 
                 Min = sapply(len_data, min),
                 Max = sapply(len_data, max),
                 Count = sapply(len_data, length))

# goseq wants a named binary vector where 1 represents DE, 0 not DE and the names are gene IDs.
gogns <- as.integer(len$Gene %in% anno.mm$ENSEMBL)
names(gogns) <- len$Gene

# weighting for each gene, depending on its length, given by the PW
# Probability Weighting Function (PWF)
# proportion of DE genes is plotted as a function of the transcript length
pwf = goseq::nullp(gogns, bias.data = len$Median) # probably looks bad because the genes weren't taken from a DE set, but from ChIP-seq, I think

GO.wall=goseq(pwf, "mm9", "ensGene")

# can be summarized in http://revigo.irb.hr/
write.table(subset(GO.wall, over_represented_pvalue < 0.05, select = c("category","over_represented_pvalue")), "GOterms.txt", quote =FALSE, row.names = FALSE, col.names = FALSE)
```

### EdgeR

```{r running_edger}
library(edgeR)
sample_info.edger <- factor(c( rep("SNF2", 5), rep("WT", 5)))
sample_info.edger <- relevel(sample_info.edger, ref = "WT")
edgeR.DGElist <- DGEList(counts = readcounts,	group = sample_info.edger)

keep <- rowSums( cpm(edgeR.DGElist) >= 1) >= 5
edgeR.DGElist <- edgeR.DGElist[keep,]
head(edgeR.DGElist$samples)
 
edgeR.DGElist <- calcNormFactors(edgeR.DGElist, method = "TMM")
edgeR.DGElist$samples

design <- model.matrix(~sample_info.edger)
edgeR.DGElist <- estimateDisp(edgeR.DGElist, design)
edger_fit <- glmFit(edgeR.DGElist, design)
edger_lrt <- glmLRT(edger_fit)

DGE.results_edgeR <- topTags(edger_lrt, n = Inf, # to retrieve all genes
							sort.by = "PValue", adjust.method = "BH")
```

### Limma-voom

```{r running_limma}
library(limma)
rownames(design) <- colnames(edgeR.DGElist)
voomTransformed <- voom(edgeR.DGElist, design, plot=FALSE)
voomed.fitted <- lmFit(voomTransformed, design = design)
voomed.fitted <- eBayes(voomed.fitted)
DGE.results_limma <- topTable(voomed.fitted, coef = "sample_info.edgerSNF2",
         number = Inf, adjust.method = "BH",
         sort.by = "logFC")

```

## Comparing the results of the different DGE tools


### How many genes were found in all tools?

__Venn diagram__:

```{r venn_plot, fig.height = 5, fig.width = 5}
library(gplots)
DE_list <- list(edger = rownames(subset(DGE.results_edgeR$table, FDR <= 0.05)),
                deseq2 = rownames(subset(DGE.results, padj <= 0.05)),
                limma = rownames(subset(DGE.results_limma, adj.P.Val <= 0.05))
)
venn(DE_list)
```

More sophisticated/useful diagram, especially if you are comparing more than 3 lists:

```{r upset_plot}
library(UpSetR)
upset(fromList(DE_list), order.by = "freq")
```

### How well do the logFC agree?

First, we need the genes that are part of all the DE lists.

```{r custom_function}
# if you want to use the following function, you will have to execute its code

#' Make data.frame of memberships
#'
#' @details This function takes as input a list of vectors (e.g., 3 lists of names)
#' and returns a data.frame that indicates for every unique entry the vector it is
#' being part of. This is the basis for all venn diagrams.
#' This function is basically the the same as \code{UpSetR::fromList()}, the only 
#' difference being that I am adding rownames to the output.
#'
#' @param in_list list
#' @return data.frame where the column names correspond to the names of the vectors
#' of the list
#' @examples
#' listInput <- list(one = c(1, 2, 3, 5, 7, 8, 11, 12, 13),
#'                  two = c(1, 2, 4, 5, 10),
#'                  three = c(1, 5, 6, 7, 8, 9, 10, 12, 13))
#' ups_input <- make_membership_df(listInput)
make_membership_df <- function(in_list){
  # get a vector of all entries
  our_universe <- unique(unlist(in_list))

  out_data <- unlist(lapply(in_list, function(x) {
    x <- as.vector(match(our_universe, x)) # NA will be introduced for every no-match
  }))

  out_data[is.na(out_data)] <- as.integer(0) # mark every non-match with a zero
  out_data[out_data != 0] <- as.integer(1) # every match gets a one
  # get the correct shape of the data.frame
  out_data <- matrix(out_data, ncol = length(in_list), byrow = F)
  out_data <- out_data[which(rowSums(out_data) != 0), ]
  out_data <- as.data.frame(out_data)
  names(out_data) <- names(in_list)
  row.names(out_data) <- our_universe
  return(out_data)
}
```

```{r eval=TRUE}
DE_gns_all <- fromList(DE_list)
rownames(DE_gns_all) <- unique(unlist(DE_list))
head(DE_gns_all)
# keep only those genes that were found to be DE in all three tools
DE_gns_all <- row.names(DE_gns_all[rowSums(DE_gns_all) == 3,])
DE_fc <- data.frame(edger = DGE.results_edgeR[DE_gns_all,]$table$logFC, 
                    limma = DGE.results_limma[DE_gns_all,]$logFC,
                    deseq2 = DGE.results[DE_gns_all,]$log2FoldChange,
                   row.names = DE_gns_all)
```

```{r pairs_logFC, fig.width = 6, fig.height = 7}
pairs(DE_fc)
```

```{r hm_logFC, fig.width = 4, fig.height = 7}
NMF::aheatmap(as.matrix(DE_fc) )
```


```{r, echo=FALSE, eval = FALSE}
save.image(file = "ReadCountsToDGE.RData")
```
