---
title: "Read counts to DGE, Part II"
output: pdf_document
---

Here, we will show you how to:

* plot correlations of replicates
* do PCA
* make a dendrogram

We assume that Part I has already been run. 

```{r setWD}
folder <- "~/Documents/Teaching/2017_RNA-seq/"
setwd(folder)
```
```{r settingUp, warning=FALSE, echo=TRUE, message=FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, message = FALSE,cache=FALSE)
library(DESeq2)
library(ggplot2)
load("featureCountExploration.RDat")
```

## Correlation plots

The ENCODE consortium recommends that _"for messenger RNA, (...) biological replicates [should] display greater than 0.9 correlation for transcripts/features"_.

The Pearson correlation coefficient is a measure of the strength of the linear relationship between two variables and is often used to assess the similarity of RNA-seq samples in a pair-wise fashion.
It is defined as the **covariance of two variables divided by the product of their standard deviation**.

```{r correlation}
# to plot the correlation with the pairs() function, we need to define
# a function that will calculate the correlation and print it as text
# (code is in fact taken from the help text of ?pairs() )
panel.cor <- function(x, y, digits = 2, prefix = "", cex.cor, ...)
{
    usr <- par("usr"); on.exit(par(usr))
    par(usr = c(0, 1, 0, 1))
    
    # define the type of correlation
    corr_coeff <- cor(x, y, method = "pearson") 
    
    txt <- format(c(corr_coeff, 0.123456789), digits = digits)[1]
    txt <- paste0(prefix, txt)
    if(missing(cex.cor)) cex.cor <- 0.8/strwidth(txt)
    text(0.5, 0.5, txt, cex = cex.cor * abs(corr_coeff))
}

# let's make the plots
pairs(log.norm.counts[,c(paste("WT",c(1:5),sep="_"))], lower.panel = panel.smooth, upper.panel = panel.cor)

pairs(log.norm.counts[,c(paste("SNF2",c(1:5),sep="_"))], lower.panel = panel.smooth, upper.panel = panel.cor)
```

## Hierarchical clustering

Hierarchical clustering requires two decisions: 

1. How should the (dis)similarity between pairs be calculated? [distance]
2. How should the (dis)similarity be used for the clustering? [linkage]

Let's start with a dendrogram determined by hierarchical clustering of a Pearson correlation distance matrix:

```{r fig.width=10, fig.height=5}
# cor() calculates the correlation between columns of a matrix
distance.m <- as.dist(1 - cor(log.norm.counts, method = "pearson"))
plot(hclust(distance.m), labels = colnames(log.norm.counts), main = "seq.depth and log2 transformed read counts\ndistance: Pearson correlation")

distance.m_rlog <- as.dist(1 - cor(assay(rlog.DESeq.sumExp), method = "pearson" ))
plot(hclust(distance.m_rlog), labels = colnames(log.norm.counts), main = "rlog transformed read counts\ndistance: Pearson correlation")
```

Which linkage method was used?

The most popular choices for the \textit{linkage function} are:

* _complete_: intercluster distance = largest distance between any 2 members of either cluster
* _average_: intercluster distance = average distance between any 2 members
* _single_:  intercluster distance = shortest distance between any 2 members

_complete_ often performs best.

Can you make a dendrogram with Euclidean distance and linkage method "average"?

```{r echo=FALSE, eval=TRUE}
# dist() computes the distances between the _rows_ of a data matrix
plot(hclust( dist( t(log.norm.counts), method = "euclidean"), method = "average"),
  labels=colnames(log.norm.counts),
  main = "seq.depth and log2 transformed read counts")

plot(hclust( dist( t(assay(rlog.DESeq.sumExp)), method = "euclidean"), method="average"),
     labels=row.names(colData(rlog.DESeq.sumExp)),
     main = "rlog transformed read counts")
```


## Principal Components Analysis

The principal components (PCA) plot is a useful diagnostic for examining relationships between samples. 
Principal components represent the directions along which the variation in the data is maximal, so that a few dimensions can be used to represent the information from thousands of mRNAs per sample.
More details can be found in the main text; here we will focus on the `R` code needed to do a PCA on normalized and transformed read count data.

PCA is performed using the function `prcomp`.

```{r}
pc <- prcomp(t(rlog.norm.counts))
str(pc)
head(pc$x)

plot(pc$x[,1], pc$x[,2], 
     col = colData(DESeq.ds)[,1], pch = as.integer(colData(DESeq.ds)[,1]),
     main = "PCA of seq.depth normalized and log2 transformed read counts")

# add legend
legend("bottomleft", pch = unique(as.integer(colData(DESeq.ds)[,1])), legend = unique(colData(DESeq.ds)[,1]) )
```

How could you focus the PCA on 1,000 genes that have the greatest variability between the samples? Hint: `var()` can be used to calculate the variation for a vector.

```{r echo=FALSE, eval=TRUE}
# gene-wise variation (= variation per row)
variation.readcounts <- apply(rlog.norm.counts, 1, FUN = "var")
# get top 1000 row indeces
topgenes <- head(order(variation.readcounts, decreasing=TRUE), 1000)
# use the indeces
pc <- prcomp(t(rlog.norm.counts[topgenes,]))
```

`DESeq2` provides a convenient function for PCA of `DESeqDataSets`.

```{r PCA_DESeq2}
print(plotPCA(DESeqTransform(DESeq.ds), ntop = 500) + 
        theme_bw() + ggtitle("Raw counts"))

print( plotPCA(rlog.DESeq.sumExp) + theme_bw() + ggtitle("Rlog transformed counts"))
```

Even more convenient and comprehensive: `pcaExplorer`, an R application with numerous functionalities.

```{r eval = FALSE}
#source("https://bioconductor.org/biocLite.R")
#biocLite("pcaExplorer")
library(pcaExplorer)
pcaExplorer(dds = DESeq.ds, rlt = DESeq.rlog)
```

```{r, echo=FALSE}
save.image(file = "ReadCountsToDGE.RData")
```
